---
title: "Student_Por"
author: "Ambar Mishra"
date: "5/23/2021"
output: html_document
---


## Background

Introduce each variable and describe the possible effects it will have on it with our predictions on what will happen

## Questions of Interest

## Analysis Overview

## Results

## Importing Dataset

Here is our imported dataset.

```{r}
student_data <- read.csv('student-por.csv')
```

# Data Cleaning

Here is our data cleaning section. Our first step of data cleaning was to convert the "yes" and "no"'s into 0's and 1's. 


## Schoolsup Column 

School sup column is extra educational support. 

```{r}
student_data$schoolsup <- ifelse(student_data$schoolsup == 'yes', 1, 0)
student_data$schoolsup
```

## Famsup column

Family sup column is for extra family support

```{r}
student_data$famsup <- ifelse(student_data$famsup == 'yes', 1, 0)
student_data$famsup
```


## Paid column

Paid column is for extra paid classes for course subjects. 

```{r}
student_data$paid <- ifelse(student_data$paid == 'yes', 1, 0)
student_data$paid
```


## Activities column

```{r}
student_data$activities <- ifelse(student_data$activities == 'yes', 1, 0)
student_data$activities
```

## Nursery Column

```{r}
student_data$nursery <- ifelse(student_data$nursery == 'yes', 1, 0)
student_data$nursery
```

## Higher Column

```{r}
student_data$higher <- ifelse(student_data$higher == 'yes', 1, 0)
student_data$higher
```

## Internet Column 

```{r}
student_data$internet <- ifelse(student_data$internet == 'yes', 1, 0)
student_data$internet
```

## Romantic Column

```{r}
student_data$romantic <- ifelse(student_data$romantic == 'yes', 1, 0)
student_data$romantic
```

Note: check a relationship between internet and grades. 

```{r}
View(student_data)
```


## Simple Linear Regression to study the relationship between absences and grades

```{r}
attach(student_data)
```

```{r}
linear_model = lm(G3~absences)
summary(linear_model)
```

Absences is clearly significant, and we can conclude that the coefficient for absences is nonzero. The regression equation from the model is G3 = 12.13880 - 0.06361*absences

Low R-squared





```{r}
plot(linear_model)
```

Points are approximately normally distributed
There seems to be 2 leverage points
Residual assumptions seem to be violated

## Logistic model with Internet and G3

```{r}
logistic_model =glm(G3~internet)
logistic_model
```

```{r}
summary(logistic_model)
```




