---
title: "Student_Por"
author: "Ambar Mishra"
date: "5/23/2021"
output: html_document
---

## Importing Dataset

Here is our imported dataset.

```{r}
student_data <- read.csv('student-por.csv')
```

# Data Cleaning

Here is our data cleaning section. Our first step of data cleaning was to convert the "yes" and "no"'s into 0's and 1's. 


## Schoolsup Column 

School sup column is extra educational support. 

```{r}
student_data$schoolsup <- ifelse(student_data$schoolsup == 'yes', 1, 0)
student_data$schoolsup
```

## Famsup column

Family sup column is for extra family support

```{r}
student_data$famsup <- ifelse(student_data$famsup == 'yes', 1, 0)
student_data$famsup
```


## Paid column

Paid column is for extra paid classes for course subjects. 

```{r}
student_data$paid <- ifelse(student_data$paid == 'yes', 1, 0)
student_data$paid
```


## Activities column

```{r}
student_data$activities <- ifelse(student_data$activities == 'yes', 1, 0)
student_data$activities
```

## Nursery Column

```{r}
student_data$nursery <- ifelse(student_data$nursery == 'yes', 1, 0)
student_data$nursery
```

## Higher Column

```{r}
student_data$higher <- ifelse(student_data$higher == 'yes', 1, 0)
student_data$higher
```

## Internet Column 

```{r}
student_data$internet <- ifelse(student_data$internet == 'yes', 1, 0)
student_data$internet
```

## Romantic Column

```{r}
student_data$romantic <- ifelse(student_data$romantic == 'yes', 1, 0)
student_data$romantic
```

Note: check a relationship between internet and grades. 
