---
title: "Student_Por"
author: "Ambar Mishra"
date: "5/23/2021"
output: html_document
---


## Background

Introduce each variable and describe the possible effects it will have on it with our predictions on what will happen

## Questions of Interest

## Analysis Overview

## Results

## Importing Dataset

Here is our imported dataset.

```{r}
student_data <- read.csv('student-por.csv')
```

# Data Cleaning

Here is our data cleaning section. Our first step of data cleaning was to convert the "yes" and "no"'s into 0's and 1's. 


## Schoolsup Column 

School sup column is extra educational support. 

```{r}
student_data$schoolsup <- ifelse(student_data$schoolsup == 'yes', 1, 0)
student_data$schoolsup
```

## Famsup column

Family sup column is for extra family support

```{r}
student_data$famsup <- ifelse(student_data$famsup == 'yes', 1, 0)
student_data$famsup
```


## Paid column

Paid column is for extra paid classes for course subjects. 

```{r}
student_data$paid <- ifelse(student_data$paid == 'yes', 1, 0)
student_data$paid
```


## Activities column

```{r}
student_data$activities <- ifelse(student_data$activities == 'yes', 1, 0)
student_data$activities
```

## Nursery Column

```{r}
student_data$nursery <- ifelse(student_data$nursery == 'yes', 1, 0)
student_data$nursery
```

## Higher Column

```{r}
student_data$higher <- ifelse(student_data$higher == 'yes', 1, 0)
student_data$higher
```

## Internet Column 

```{r}
student_data$internet <- ifelse(student_data$internet == 'yes', 1, 0)
student_data$internet
```

## Romantic Column

```{r}
student_data$romantic <- ifelse(student_data$romantic == 'yes', 1, 0)
student_data$romantic
```

Note: check a relationship between internet and grades. 

```{r}
View(student_data)
```


## Simple Linear Regression to study the relationship between absences and grades

```{r}
attach(student_data)
```

```{r}
linear_model = lm(G3~absences)
summary(linear_model)
```

Absences is clearly significant, and we can conclude that the coefficient for absences is nonzero. The regression equation from the model is G3 = 12.13880 - 0.06361*absences

Low R-squared





```{r}
plot(linear_model)
```

We can see that in the normal qq plot, there is a problem with the left tail, since many points are not on the dotted dashed line. Clearly, some data points are affecting the normal distribution assumption, and we hypothesize that it is the data points with 0 as the grade. We can take a look at the number of people who receive a 0 as their grade.

```{r}
nrow(student_data[G3 == 0,])
```


15 of the grades are considered to be zeroes, which clearly affects the distribution of the grades. We can see these 15 points on the normal qq plot that greatly affects the left tail.

```{r}
hist(G3)
```

We can also see this in the histogram, where without the zeroes, the data for the final grade seems approximately normally distributed.

We can assume that the grades were either misreported or if the students received a 0 for a particular reason. Either way, these students are outliers in the dataset. We shall remove the outliers with a grade of 0 to see if our model fit and assumptions will be better. 

```{r}
student_data_no_zeroes = student_data[-c(which(G3 == 0)),]
```

Here is the new dataset we will be working with that contains no zeroes for the final grade. We now will run the linear model we did previoulsy and compare the two models. 

```{r}
attach(student_data_no_zeroes)
```

```{r}
linear_model_2 = lm(G3~absences)
summary(linear_model_2)
```

We can see that the coefficient for absences has become more significant with a p-value of 6.54e-07, which suggests that there is enough evidence for the coefficient of absences to be non-zero. The F-statistics also increased as well as the multiple r squared.

```{r}
hist(G3)
```

Clearly, the histogram of the final grade is much better than before.

```{r}
plot(linear_model_2)
```

We can see that after removing the outliers, the assumption of normality is met. In addition, for the residual plot the assumption of constant error variance and independence is met. Since removing the students who received a 0 for their final grade are clearly outliers and affect our model greatly, we will continue with the subset of the data without zeroes in the grades when adding other variables into the model.





## Logistic model with Internet and G3

```{r}
logistic_model =glm(internet~G3, family = "binomial")
logistic_model
```

```{r}
summary(logistic_model)
```

```{r}
prob = predict(logistic_model, type = "response")
```




